NCBI

### 2025-04-01 episodes/ncbi.Rmd
First half of NCBI redone with more hints and solutions

Exercise 1
- How do you efficiently and reliably navigate NCBI's website?
Exercise 2
- How do you search sequences?
- How do you use the right BLAST flavor?

@@

Exercise 1
- Familiarize yourself with the differences NCBI databases
- Browse and search the databases, using cross-links
Exercise 2
- Use and understand blast alignment tools
- Characterize sequences
- Understand genomic context and synteny

@@

This part is to encourage you to explore NCBI resources. Questions are examples or real-life questions that you might ask yourself later. There are not necessarily exactly one solution to the question. 
Start by going to NCBI's web site, main search page: https://www.ncbi.nlm.nih.gov/search/. There, you have the list of most of NCBI’s databases, sorted by category. Take some time to explore the following sections: Literature, Genomes, Genes and Proteins.

@@

1. an article about *Escherichia coli* O104:H4 published by Matthew K Waldor in the New England Journal of Medicine; and 
1. a paper about *E. coli* whose last author is L. Wang, published in PLoS ONE in 2014.

@@

Use the advanced search from PubMed, by clicking on “Advanced” (tutorial https://www.ncbi.nlm.nih.gov/books/NBK3827/#pubmedhelp.PubMed_Quick_Start), to build an exact search. Use the different fields to build a search that returns only a single match.

@@

1. Rasko DA, Webster DR, […], Waldor MK. Origins of the *E. coli* strain causing an outbreak of hemolytic-uremic syndrome in Germany. N Engl J Med. 2011 Aug 25;365(8):709-17. doi: 10.1056/NEJMoa1106920. Epub 2011 Jul 27.
1. Zhang X, Li Y, Liu B, Wang J, Feng C, Gao M, Wang L. Prevalence of veterinary antibiotics and antibiotic-resistant *Escherichia coli* in the surface water of a livestock production region in northern China. PLoS One. 2014 Nov 5;9(11):e111026. doi: 10.1371/journal.pone.0111026. eCollection 2014.

@@

Now try to find the most recent paper citing the first article by Rasko et al above. NCBI provides the “Cited by” link (right menu), but you can also use

- Google Scholar (https://scholar.google.com) or 
- Web of Knowledge (https://www.webofknowledge.com; Access from UU: https://www.ub.uu.se/ > Databases A-Z > W > Web of Science Core Collection). The language settings of Google Scholar can be adjusted in the menu under Settings.

@@

Nucleotide database contains mostly **genome** records, so the Shiga toxin gene might be anywhere there.

@@

@@

@@

All the information in that section ("Related information") is coming through the Entrez system. In the `FASTA` record, look for `\dbxref` (database cross-reference) which provides the Entrez links.

@@

You can change the format by clicking on the top left corner.
Look for "Related information" in the right menu.

@@

Can you find just the sequence information in `FASTA` format? How is the format organized? What is included in the header? Discuss the advantages of the two formats. Can you find a definition of both formats on the NCBI website or elsewhere?

@@

@@

Can you display the genome record on which this gene is encoded?

>>>>

Exercise 1
- How do you efficiently and reliably navigate NCBI's website?

Exercise 2
- How do you search sequences?
- How do you use the right BLAST flavor?

@@

Exercise 1
- Familiarize yourself with the differences NCBI databases
- Browse and search the databases, using cross-links

Exercise 2
- Use and understand blast alignment tools
- Characterize sequences
- Understand genomic context and synteny

@@

This part is to encourage you to explore NCBI resources. Questions are examples or real-life questions that you might ask yourself later. There are not necessarily exactly one solution to the question.

Start by going to NCBI's web site, main search page: https://www.ncbi.nlm.nih.gov/search/. There, you have the list of most of NCBI’s databases, sorted by category. Take some time to explore the following sections: Literature, Genomes, Genes and Proteins.

@@

1. an article about *Escherichia coli* O104:H4 published by Matthew K Waldor in the New England Journal of Medicine; and 
1. a paper that has *Escherichia coli* in the title, whose last author is L. Wang, and was published in PLoS ONE in 2014.

@@

Use the advanced search from PubMed, by clicking on “Advanced” (tutorial https://www.ncbi.nlm.nih.gov/books/NBK3827/#pubmedhelp.PubMed_Quick_Start), to build an exact search. Use the different fields to build a search that returns only a single match. In particular, look at the [Search field tags](https://pubmed.ncbi.nlm.nih.gov/help/#using-search-field-tags).

Use search terms with search tags ("search term"[searchTag]")and combine them with Boolean operators ("AND").

Useful search tags: Author [au], Date of publication [dp], Journal [ta], Last author [lastau], Title [ti].

@@

1. Waldor[au] AND "Escherichia coli O104:H4" AND NEJM[ta]
1. "Wang L"[lastau] AND "Escherichia coli"[ti] AND 2014[dp] and "PLOS One"[ta].

@@

Now try to find the most recent paper citing the first article by Rasko et al above. Compare the use of

- NCBI (“Cited by” link) and
- Google Scholar (https://scholar.google.com) (the language settings of Google Scholar can be adjusted in the menu under Settings) and
- Web of Knowledge (https://www.webofknowledge.com; Access from UU: https://www.ub.uu.se/ > Databases A-Z > W > Web of Science Core Collection).

@@

Nucleotide database contains mostly **genome** records, so the Shiga toxin gene might be anywhere on these contigs/genomes.

@@

There is often a "Bibliography" section. There is also often a "PubMed" link under "Related Information" in the right menu.

@@

The taxonomic information is under the "Summary" section, under "Organism" and "Lineage".

@@

All the information in that section ("Related information") is coming through the Entrez system. In `GenBank` records, look for `\dbxref` (database cross-reference) which provides the Entrez links.

@@

You can change the format by clicking on the top left corner.

Look also for "Related information" in the right menu and for the `/db_xref` links in the main file.

@@

Can you find only the sequence information in `FASTA` format? How is the format organized? What is included in the header? Discuss the advantages of the two formats. Can you find a definition of both formats on the NCBI website or elsewhere?

@@

Not all proteins have 3D structures. If your protein doesn't have one, try the following Shiga toxin:

https://www.ncbi.nlm.nih.gov/protein/NP_311001.1

In the right menu, under "Related Information", there may be a "Related Structures" link.

See here: https://www.ncbi.nlm.nih.gov/structure?LinkName=protein_structure_related_list&from_uid=15832228

@@

Can you display the genome record on which the gene is encoded?

If you lost the gene page, here is another one:

https://www.ncbi.nlm.nih.gov/gene/26516284

There are several ways to display the GenBank file. 
- In the "Genomic regions, transcripts, and products" section, there is a link to the GenBank record
- There is another one in the "Related sequences" section (column "Accession and Version").
